---
title: "I-SPY Project Data Processing"
author: "Nasim Bondar Sahebi and Katy Scott"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    theme: lumen
    highlight: espresso
    code_folding: show
  rmdformats: default
affiliation: Princess Margaret Cancer Center, University Health Network, Toronto,
  Canada
params:
  palette: npg
  size: 1
  showlabels: false
  ellipse: true
css: styles/custom.css
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---
## Overview

This Script **processes and prepares** the **I-SPY2 dataset**, focusing on:
- **Cleaning, merging, and structuring** genomics and radiomics clinical data.
- **Identifying common patients** between datasets.
- **Ensuring MRI segmentation consistency** for further analysis.

### Dataset Summary
- **Study Reference**: [PMID: 35623341](https://pubmed.ncbi.nlm.nih.gov/35623341/)  
- **Clinical Trial**: [ISPY2/NCT01042379](https://clinicaltrials.gov/study/NCT01042379)  
- **Data Source**: [GEO: GSE194040](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE194040)  
- **Study Population**:  
  - **Genomics Data**: 987 patients (mRNA & RPPA profiling)  
  - **Radiomics Data**: 985 patients (MRI scans)  
- **Platforms Used**:
  - **Genomics**: Agilent 44K arrays  
  - **Radiomics**: MRI (DCE-MRI)   

### Patient and Treatment Details
- **Total Patients (RNA-seq)**: 987  
- **Total Patients (MRI)**: 985 (719 I-SPY2 cases, 266 ACRIN-6698 cases)  
- **Treatment Groups**:
  - **IO + Chemo**: 69 patients  
  - **Chemo + Targeted Therapy**: 687 patients  
  - **Chemo Alone**: 179 patients  
  - **Targeted Therapy Alone**: 52 patients  
- **Key Treatment Arms**:
  - Paclitaxel ± Trastuzumab  
  - Veliparib + Carboplatin  
  - AMG-386 (Trebananib) ± Trastuzumab  
  - Neratinib, Ganetespib, Ganitumab, Pembrolizumab, Pertuzumab, T-DM1  
  
#### Treatment Arms
- **Chemotherapy ± IO**  
  - Paclitaxel ± Trastuzumab  
  - Veliparib + Carboplatin  
- **Targeted Therapies**  
  - AMG-386 (Trebananib) ± Trastuzumab  
  - Neratinib, Ganetespib, Ganitumab, Pembrolizumab, Pertuzumab, T-DM1
  
### Additional Information
- **Batch Correction**:  
  - **ComBat** was applied to correct batch effects in gene expression data.  
  - Adjusted ~800 patients; linear adjustments applied to remaining cases.  

- **Radiomics Data**:  
  - MRI dataset includes **4 serial DCE-MRI scans per patient**.  
  - Performed **before and during neoadjuvant chemotherapy** to track tumor volume changes and response.  

- **Data Availability**:  
  - 📂 **Genomics**: Available in [GEO: GSE194040](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE194040).  
  - 📂 **Imaging**: MRI datasets accessible via [The Cancer Imaging Archive (ISPY2 Collection)](https://www.cancerimagingarchive.net/collection/ispy2/).  

- **Key Reference**:  
  - Wolf DM et al., *Cancer Cell*, 2022. [PMID: 35623341](https://pubmed.ncbi.nlm.nih.gov/35623341).  

## Data Processing Steps

### Load Required Libraries
 
```{r libraries,include=TRUE,message=FALSE,warning=FALSE}
library(here)
library(MultiAssayExperiment) 
library(knitr)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggvenn)
library(kableExtra) 
```

### Load & Extract Data {.tabset .tabset-pills}

Load MultiAssayExperiment `.rds` File**
This step extracts **clinical, expression(Microarray quantile normalizated), and annotation data** for further analysis.

```{r load multiassay and extract Data}
# Define base directory.
base_dir <- file.path(here(), "workflow")

# Load dataset
mae <- readRDS(file.path(base_dir, "Source Data/ICB_Wolf.rds"))

# Extract clinical, expression, and annotation data
clin <- data.frame(colData(mae))  
expr <- assays(mae)[["expr"]]     
annot <- data.frame(rowData(mae@ExperimentList$expr))
```

#### Clinical Data 

Display first 50 rows of  Clinical Data.

```{r clin summery}
# Display clin 
kable(head(clin,50), caption = "Genomics Clinical Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Microarray Expression Data 

Display first 20 rows of  quantile-normalized Microarray data

```{r expr summery}
# Display the first 20 rows 
kable(head(expr, 20), caption = "Expression Data-Microarray") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Annotation Data 

Displays gene annotation for the dataset.

```{r data summery}
# Display the first 20 rows 
kable(head(annot, 20), caption = "Gene Annotation Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Match Patients: Genomics & Radiomics Integration

This section identifies overlapping patients between genomics (gene expression) and radiomics (MRI scans) datasets.

#### Step 1: Load Radiomics Metadata

The radiomics dataset contains dynamic contrast-enhanced MRI **(DCE-MRI)** scans used to monitor tumor response to treatment.  

**Metadata Source:**  
- **File:** `ISPY2-Imaging-Cohort-1-Clinical-Data.xlsx`  
- **Access:** [The Cancer Imaging Archive - I-SPY2 Collection](https://www.cancerimagingarchive.net/collection/ispy2/)  
- **Total Patients:** **985**  
  - **719** I-SPY2 cases  
  - **266** [ACRIN-6698 cases](https://www.cancerimagingarchive.net/collection/ispy1/)  

```{r Load Radiomics Metadata}
# Load radiomics clinical data
Radiomics_clin <- read_excel(file.path(base_dir, "Source Data/ISPY2-Imaging-Cohort-1-Clinical-Data.xlsx"))  

# Display radiomics metadata with scrolling
kable(Radiomics_clin, caption = "Radiomics Clinical Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Step 2: Summary of Radiomics Data

- **Each patient has 4 MRI scans (T1-T4)**
- **MRI features include:**
  - Functional Tumor Volume (FTV)
  - Longest Tumor Diameter
  - Sphericity  
- **Data collected from 22+ clinical centers**   

#### Step 3: Identify Common Patients

To integrate genomics and radiomics data, we **match patient IDs** across datasets.

```{r check Patient Matching}
# Prepare genomic clinical data
Genomics_clin <- clin  
Genomics_clin$patientid <- sub("^X", "", Genomics_clin$patientid)  # Remove "X" prefix
rownames(Genomics_clin) <- Genomics_clin$patientid
colnames(expr) <- sub("^X", "", colnames(expr))

# Identify common patients
common_patients <- intersect(Genomics_clin$patientid, Radiomics_clin$Patient_ID)  

# Print count of common patients
cat("Patients with both microarray and MRI data:", length(common_patients), "\n")
```

#### Step 4: Visualize Patient Overlap

A Venn diagram illustrates genomics vs
. radiomics patient overlap.

```{r visualize patient overlap}
# Create Venn diagram
venn_data <- list(Genomics = Genomics_clin$patientid, Radiomics = Radiomics_clin$Patient_ID)

ggvenn(venn_data, fill_color = c("#A6CEE3", "#FB9A99"), stroke_size = 0.5, set_name_size = 4) +
  ggtitle("Overlap of Genomics (Microarray) and Radiomics (MRI) Patient IDs") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), plot.background = element_rect(fill = "white"), panel.background = element_blank())

```

## Merge Genomics & Radiomics Data

Merging both datasets based on **common patient IDs**.

**Adjustments:**  
- **Prefix MRI columns with `"Radiomics_"`**  
- **Ensure consistent patient IDs**  
- **Save merged dataset for analysis**  
 
```{r Merge}
# Add prefix to radiomics columns
colnames(Radiomics_clin)[-which(colnames(Radiomics_clin) == "Patient_ID")] <- paste0("Radiomics_", colnames(Radiomics_clin)[-which(colnames(Radiomics_clin) == "Patient_ID")])

# Merge datasets
merged_clin <- merge(Genomics_clin, Radiomics_clin, by.x = "patientid", by.y = "Patient_ID")

# Sort by patient ID
merged_clin <- merged_clin[order(merged_clin$patientid), ]

# Display merged dataset with scrolling
kable(head(merged_clin,50), caption = "Merged Clinical Data (Genomics & Radiomics)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

# Save dataset as CSV
# write.csv(merged_clin, "~/BHK lab/I-SPY-Project/output Data/Merged_clin_GenomicandRadiomic.csv", row.names = FALSE)
```

### Segmentation Consistency Check

To ensure accurate **Functional Tumor Volume (FTV)** analysis in the I-SPY clinical trials, we verify that each patient has the expected **4 segmentation masks** from **Dynamic Contrast Enhanced (DCE) MRI** scans.  

### Segmentation Process in DCE-MRI

Each scan undergoes multiple segmentation steps:  
1. **Manual VOI (Volume of Interest):** A rectangular region around the enhancing tumor.  
2. **Background Masking:** Removes noise and fat saturation artifacts.  
3. **Percent Enhancement (PE) Threshold:** Filters non-enhancing tissue.  
4. **Connectivity Filter:** Ensures 3D tumor voxel connectivity.  
5. **OMIT Regions:** Manual exclusion of non-tumor areas.  

**Segmentation masks are stored as 2D DICOM images** aligned with the original **DCE MRI scans**.  
note: **Download masks from UCSF Box:** [Link](https://ucsf.box.com/s/d09xetaatzgnz2p9bxlqh50rrzilgcoj)  


### Segmentation Metadata {.tabset .tabset-pills}

#### ISPY1: MRI Series Metadata  
**Metadata for ISPY1** is stored in `ispy1_dce_series_info.xlsx`, containing data for **221 patients**. This dataset includes MRI series metadata for each patient, covering:  

- **Patient ID & Time Point (T1–T4)**  
- **MRI Series Description & Unique ID**  
- **Signal Enhancement Ratio (SER) Calculation Indicator**  
- **OMIT Region Counts (Non-Tumor Exclusion)**   

```{r ISPY1 segmentation metadata}
# Load ISPY1 segmentation metadata
ispy1_dce_series_info <- read_excel(
  path = file.path(base_dir, "Source Data", "ispy1_dce_series_info.xlsx"), 
  sheet = "DCE Series, All 20160823"
) 

# Display ISPY1 segmentation metadata
kable(head(ispy1_dce_series_info, 20), caption = "DCE MRI Series Information") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### ISPY2: MRI Series Metadata  

Metadata for ISPY2 was extracted from `series-data1741883133411.csv`, containing data for **719 patients**. This dataset includes MRI series metadata for each patient, covering:  

- **Subject ID**  
- **Study Description** (e.g., `"ISPY2_MRI_T0"`)  
- **Number of Images**  
- **Modality** (e.g., `"MR"`, `"SEG"`)  

```{r segmentation metadata}
# Load ISPY2 segmentation metadata
ispy2_dce_series_info <- read.csv(file.path(base_dir, "Source Data", "series-data1741883133411.csv"))

# Rename the "ispy2_dce_series_info" column to "PATIENT_ID"
colnames(ispy2_dce_series_info)[colnames(ispy2_dce_series_info) == "Subject.ID" ] <- "PATIENT_ID"

# Extract time points from the "Study.Description" column
ispy2_dce_series_info$TIME_POINT <- gsub(".*(T[0-9]+).*", "\\1", ispy2_dce_series_info$Study.Description)

# Display ISPY2 segmentation metadata
kable(head(ispy2_dce_series_info, 20), caption = "ISPY2 DCE MRI Series Information") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

# Select only PATIENT_ID and TIME_POINT columns from both datasets
ispy1_selected <- ispy1_dce_series_info[, c("PATIENT_ID", "TIME_POINT")]
ispy2_selected <- ispy2_dce_series_info[, c("PATIENT_ID", "TIME_POINT")]

# Merge based on PATIENT_ID and TIME_POINT
ispy1and2 <- merge(ispy1_selected, ispy2_selected, by = c("PATIENT_ID", "TIME_POINT"), all = TRUE)
```

### Check Segmentation Consistency Across Time Points (T1-T4)  

Each patient (Total: 940) should have at least four segmentation masks, corresponding to their four MRI scans.

```{r Check Segmentation}
# Count segmentations per patient for each time point
series_per_patient <- ispy1and2 %>%
  group_by(PATIENT_ID, TIME_POINT) %>%
  summarize(Series_Count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = TIME_POINT, values_from = Series_Count, values_fill = 0) %>%
  mutate(across(starts_with("T"), as.numeric),
         Total_Segmentations = rowSums(select(., starts_with("T")), na.rm = TRUE))

# Display segmentation counts per patient
kable(series_per_patient, caption = "Segmentation Counts per Patient") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Patients with Missing Segmentations

Patients missing **one or more segmentation masks** need further review (24 patients).

```{r missing segmentations}
# Identify patients with <4 segmentations
patients_less_than_4 <- series_per_patient %>%
  filter(Total_Segmentations < 4) %>%
  select(PATIENT_ID, Total_Segmentations)

# Display patients missing segmentations
kable(patients_less_than_4, caption = "Patients with Less Than 4 Segmentation Masks") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```
